Provisional Title

### 2021-04-21 draft/draft.tex
draft skeleton

\synctex=1
\documentclass[a4paper,10pt]{article}

%% TO-DO: Achim and Schill in acknowl
%% TO-DO: link to repo in paper and suppl mat
%% TO-DO: my new grant
%% TO-DO: give URLs to, e.g, Crooks reference: mybibwithurl?

%% For tracking changes
%% https://tex.stackexchange.com/questions/65453/track-changes-in-latex

% packages

% %% txfonts breaks amsmath.
% %% http://www.texfaq.org/FAQ-alreadydef

\usepackage{gitinfo2}

\usepackage{savesym}
\usepackage{amsmath}
\savesymbol{iint}
\usepackage{txfonts}
\restoresymbol{TXF}{iint}

\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
%\usepackage{txfonts} % nice fonts
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{subcaption} % collages of multiple images
%%\usepackage[hidelinks]{hyperref} % hyperlinks
\usepackage{hyperref}
% \usepackage{cite} % contraction of references, use
%                   % \usepackage[superscript]{cite} for superscript
%                   % citations
%% cite does not follow the usual pre and post citation
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage{xcolor} % text color
\usepackage[margin=3cm]{geometry} % margins
\usepackage{lineno} % line numbers
\usepackage{csquotes} % paragraph quotes
\usepackage{multirow} % joining rows in tables
\usepackage{array}
\usepackage[normalem]{ulem}
\usepackage{todonotes}
%%\usepackage{amsmath}: something breaks. What?
\usepackage{pdflscape} %% Landscape
\usepackage{ulem} % strike-through text

%% cross references with suppl mat
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\zxrsetup{toltxlabel}
\zexternaldocument*[S-]{supp}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan %% controls color of external ref, if used
}

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default}, labelsep=period, name={Fig.}} % custom figure captions
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\change}[2]{\textcolor{red}{\sout{#1}}\textcolor{blue}{#2}}
\linenumbers % add line numbers to whole doc

% title, authors, affiliations
\title{Conditional prediction of consecutive tumor evolution using cancer
  progression models: What genotype comes next?}
%\title{Conditional prediction of short-term tumor evolution using cancer progression models}
%\title{Prediction of consecutive tumor evolution using cancer progression models}
%\title{Conditional prediction of upcoming tumor evolution using cancer progression models}
%\title{Partial prediction of tumor evolutionary trajectories using cancer progression models}
%\title{Using cancer progression models to predict upcoming tumor evolution}
\author[1,2,3]{Juan Diaz-Colunga}
\author[1,2,$\dagger$]{Ramon Diaz-Uriarte}
\affil[1]{Dpt. of Biochemistry, School of Medicine, Universidad Autónoma de Madrid, Madrid, Spain}
\affil[2]{Instituto de Investigaciones Biomédicas `Alberto Sols'
  (UAM-CSIC), Madrid, Spain}
\affil[3]{Department of Ecology \& Evolutionary Biology and Microbial
  Sciences Institute, Yale University, New Haven, CT, USA}
%\affil[3]{Dpt. of Ecology and Evolutionary Biology, Yale University, New Haven, Connecticut, USA}
%\affil[4]{Microbial Sciences Institute, Yale University, New Haven, Connecticut, USA}
\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont ramon.diaz@iib.uam.es}
%\date{}

@@

% \date{\gitAuthorDate\ {\footnotesize (Release\gitRels: Rev:
%      \gitAbbrevHash)}}

@@

  Accurate prediction of tumor progression is key for adaptive
  therapy and precision medicine. Cancer progression models (CPMs) can be
  used to infer dependencies in mutation accumulation from cross-sectional
  data and provide predictions of tumor progression paths. But their
  performance when predicting the complete evolutionary paths is limited
  by violations of assumptions and the size of available data
  sets. Instead of predicting full tumor progression paths, we can focus
  on short-term predictions, more relevant for diagnostic and therapeutic
  purposes. Here we examine if five distinct CPMs can be used to answer
  the question ``Given that a genotype with $\mathrm{n}$ mutations has
  been observed, what genotype with $\mathrm{n}+1$ mutations is next in
  the path of tumor progression'' or, shortly, ``What genotype comes
  next''. Using simulated data we find that under specific combinations of
  genotype and fitness landscape characteristics CPMs can provide
  predictions of short-term evolution that closely match the true
  probabilities, and that some genotype characteristics (fitness and
  probability of being a local fitness maximum) can be much more relevant
  than global features. Thus, CPMs can provide short-term predictions even
  when global, long-term predictions are not possible because fitness
  landscape- and evolutionary model-specific assumptions are violated.
  When good performance is possible, we observe significant variation in
  the quality of predictions of different methods. Genotype-specific and
  global fitness landscape characteristics are required to determine which
  method provides best results in each case.  Application of these methods
  to 25 cancer data sets shows that their use is hampered by lack of the
  information needed to make principled decisions about method choice and
  what predictions to trust. Fruitful use of these methods for short-term
  predictions requires adapting method's use to local genotype
  characteristics and obtaining reliable indicators of performance; it
  will also be necessary to clarify the interpretation of the method's results
  when key assumptions do not hold.

@@

%% bioniformatics:  A maximum of 150 words is recommended. https://academic.oup.com/bioinformatics/pages/instructions_for_authors
%% PLoS Comp Biol: I think around 300

%% Length of ms: For bioinformatics: Original papers: up to 7 pages, this
%% is approx. 5000 words, not including figures,
%%
%% Do as pdftotext draft.pdf - | wc -w
%% Maybe rm refs and figures before word counting
%% Place \end{document} before refs; uncomment it.

%% As of now, this is twice as long as it should be for Bioinformatics:
%% 9400 words withouth figures.

\section{Introduction}\label{sec:intro}

Accurately predicting tumor evolution would allow for better prognosis and
treatment. Unfortunately, the stochastic components in key factors that
govern tumor progression (mutation, genetic drift or clonal selection
\cite{greaves_evolutionary_2015,lipinski_cancer_2016,williams2018}, as
well as non-genetic variability \cite{Brock2009} and the tumor
microenvironment \cite{Albini2007}) fundamentally limit
predictability. This is also common to other problems
\cite{lassig_predicting_2017, losos2018} such as the evolution of
antibiotic resistance \cite{toprak2012} or virulence \cite{Day2004}, where
even small increases in predictive power could be of critical importance
for the design of treatment strategies \cite{palmer2013}.

% what are cancer progression models?
Cross-sectional high-throughput data are widely available for many cancer
types, and Cancer Progression Models (CPMs) \cite{Beerenwinkel2014,
  beerenwinkel_computational_2016}, originally developed to identify
restrictions in the order of accumulation of mutations in tumors
using cross-sectional data, could thus be key tools for predicting cancer
progression. CPMs (implicitly) encode all the possible mutational
trajectories that are compatible with the restrictions and, for some
methods, provide probabilities of different paths (figure
\ref{fig:fig1}A-B). However their performance to infer the complete
evolutionary trajectories, from the root genotype with no driver mutations
to a final, fixated genotype, is limited
\cite{diaz-uriarte2019a,hosseini2019a}; this has been attributed mainly to
violations of the underlying assumptions of the models (fitness landscapes
with a single fitness maximum in the genotype with all loci mutated,
strong selection week mutation regime \cite{gillespie1983}) and the
difficulty to acquire
large, unbiased data sets to power them \cite{diaz-uriarte2019a}.

Predicting the complete evolutionary path, however, might not be the most
relevant objective. Detection of a specific tumor composition at a given
point in time automatically discards all mutational pathways that do not
visit the observed state, reducing the amount of possible evolutionary
trajectories that can be followed from then on and thus potentially
facilitating the task of predicting further tumor progression. In this
work we address the question of how accurately CPMs can predict short-term
cancer evolution after a genotype is detected: what is the state that
comes consecutively after or ``what genotype comes next''. Even if the
predictability of the complete path is low because of violations of method
assumptions, genotype-specific features could lead to local improvements
in short-term predictability \cite{bank_predictability_2016,
  visser2018,ferretti2018}. Moreover, in contrast to the more general,
non-local descriptors, these genotype-specific factors can provide more
detailed insight regarding the requirements for the methods to be
reliable. In fact, conditional predictions on the observed genotype are
arguably much more relevant than complete evolutionary path predictions
for both adaptive therapy \cite[e.g.,][]{melnikov2020,hansen2020a,
  stankova2019a} and for SMART (sequential, multiple assignment,
randomized trial) designs in dynamic treatment regimes for personalized
medicine \cite{tsiatis2020,chakraborty2013}.

[Fig.~\ref{fig:fig1} about here]

% Rather than analyzing full evolutionary
% trajectories, as has been done in other work
% \cite{diaz-uriarte2019a,hosseini2019a}, we focus here on the question
% ``what genotype comes next'', the partial paths from an observed tumor
% state to the state that comes consecutively after.

% This might, on the one hand, lead to improvements in predictive power,
% even when the predictability of the complete path is low because of
% violations of method assumptions: genotype-specific features could lead to
% local improvements in short-term predictability
% \cite{bank_predictability_2016, visser2018,ferretti2018}. On the other
% hand, conditional predictions on the observed genotype are much more
% relevant clinically and for treatment purposes in adaptive therapy
% \cite[e.g.,][]{melnikov2020,hansen2020a, stankova2019a} and for SMART
% (sequential, multiple assignment, randomized trial) designs in dynamic
% treatment regimes for personalized medicine
% \cite{tsiatis2020,chakraborty2013}.

% , accurately predicting the next genotype could be among
% the tools for the physician to become, in a game theory context, the
% ``rational leader'' that anticipates and steers to limit cancer's
% resistance strategies \cite{stankova2019a}; moreover, observed genotypes
% and predicted next genotypes could be of use in SMART (sequential,
% multiple assignment, randomized trial) designs in dynamic treatment
% regimes for personalized medicine \cite{tsiatis2020,chakraborty2013}.

% \cite[prenote][posnote]{hansen2020a}

% accurately predicting the
% next genotype would play a key role in adaptive therapy
% \cite[e.g.,][]{melnikov2020, hansen2020a,stankova2019a}, as ``(...) the
% ability to anticipate subsequent cancer cell responses, provides a
% critical opportunity to obtain more favorable outcomes by steering and/or
% limiting the cancer cell’s resistance strategies'' \cite[]

% are used in adaptive therapy \cite{whatever}

% form basis of SMART designs in individualized therapy
% could be used in SMART designs \cite{whatever---los dos libros}
% \cite{tsiatis2020,chakraborty2013}
% (yo: check los dos zotero dirs, and look for conditional on genotypes)

% \cite{melnikov2020}: inhibiting growth in bacteria

% \cite{hansen2020a}: adaptive therapy and cpompetitive supression: gral ref

% \cite{gluzman2020}: conditional treatment

% \cite{west2019}

% Motivated by the idea that information on the prospective advance of the
% disease, even if limited, can be of critical importance for the design of
% treatment strategies and can improve therapeutic outcomes
% \cite[p.~243i]{palmer2013}

% Information on the prospective advance of the disease, even if limited,
% can be of critical importance for the design of treatment strategies
% \idea{references, as in paper with Claudia}. \todo{YES! DO!}
% \idea{[Relate predicting next genotype to adaptive therapy ---tarea para
%   Ramon]}.

% Tumor progression is governed by a combination of factors such as mutation, genetic drift or clonal selection
% \cite{greaves_evolutionary_20152015,lipinski_cancer_2016,williams2018}, as well as non-genetic
% variability \cite{Brock2009} and the tumor microenvironment
% \cite{Albini2007}. These have important stochastic components that
% fundamentally limit predictability. The ability
% to understand and predict evolutionary dynamics in cancer would allow for
% better prognosis and treatment. Investigating the mutational profiles of
% evolving clones within a tumor is possible by means of high-throughput
% genome sequencing, but the use of accurate models of tumor evolution is
% still required to unveil the true paths of cancer progression
% \cite{McPherson2017}. This is not of exclusive relevance to cancer but
% also to problems such as the evolution of antibiotic resistance
% \cite{toprak2012} or virulence \cite{Day2004}.

% % what are cancer progression models?
% Cancer Progression Models (CPMs) identify restrictions in the order of
% accumulation of mutations within a tumor using cross-sectional data. They
% encode all the possible mutational trajectories that are compatible with
% the restrictions (figure \ref{fig:fig1}A-B), which can serve to narrow the
% search for the true distribution of tumor progression pathways. However
% their performance to infer these complete evolutionary trajectories, from
% the root genotype with no driver mutations to a final, fixated genotype,
% is limited \cite{diaz-uriarte2019a,hosseini2019a}; this has been
% attributed to various factors, such as violations of the underlying
% assumptions of the models and the difficulty to acquire large, unbiased
% data sets to power them \cite{diaz-uriarte2019a}.

% what are we studying?
% In practice, cancer can only be first detected once it has already
% advanced to some extent.

% Detection of a specific tumor composition at a given point in time
% automatically discards all mutational pathways that do not visit the
% observed state, reducing the amount of possible evolutionary trajectories
% that can be followed from then on and thus potentially facilitating the
% task of predicting further tumor progression.\todo{is it here where we
%   explain how this differs from work with Claudia?} In this work we
% address the question of how accurately CPMs can predict short-term cancer
% evolution after detection. Rather than analyzing full evolutionary
% trajectories, we focus on the partial paths from an observed tumor state
% to the state that comes consecutively after.

The question of what genotype comes next, or what is the state that comes
consecutively after, requires that we carefully define the meaning of
``consecutive'' states in this context. If cancer progresses under
conditions of strong selection and weak mutation (SSWM)
\cite{gillespie1983}, only evolutionary trajectories of increasing fitness
are possible, and the evolution of a new, higher-fitness clone results in
rapid expansion and effective exclusion of previous ones before any new
mutations take place. In this situation, there is a well defined sequence
of genotypes with progressively increasing fitness in the path to tumor
fixation, with the most abundant clone at every time being a direct
descendant of a previously most abundant one itself (figure
\ref{fig:fig1}C). But, in cancer, large population sizes and high mutation
rates can prevent successful clonal sweeps \cite{nichol2019a, zhao2016} so
that instead of SSWM, clonal interference and stochastic tunneling
\cite{devisser_empirical_2014,szendro_predictability_2013,sniegowski_beneficial_2010,
  wodarz_dynamics_2014} become common: multiple clones can coexist at
significant fractions for prolonged periods of time, and new
higher-fitness clones can evolve from low fitness parents that represented
a minority of the population
(Fig~\ref{fig:fig1}D). % As a consequence, genotypes with a large
% number of mutations %high mutational load
% can already be present at the time of sampling yet go undetected if their
% relative abundance is small (figure
% \ref{fig:fig1}D). % Deviations from SSWM
% are known to undermine the predictive power of CPMs
% \cite{diaz-uriarte2018,diaz-uriarte2019a}, as the model underlying most
% CPMs seems to assume SSWM (see Discussion). This assumption, however, is
% not always stated
% explicitly.
% , but it is explicit, for example, in Hosseini et al.'s
% \cite{hosseini2019a} use of CPMs for predicting tumor evolution (see
% \nameref{discussion}).

To formulate the question of ``what genotype comes next'' 
%To accurately raise a question that does not rely on SSWM to be well
%defined,
so that it does not depend on the SSWM assumption we designate an
\textit{observed} genotype as that which is most abundant in the tumor at
the time of sampling. On the other hand, a Line of Descent (LOD) is
defined as a sequence of parent-child genotypes from the wild-type to the
(possibly local) 
fitness maximum where fixation occurs \cite{szendro_predictability_2013}.
By construction, assuming that mutations accumulate one by one and cannot
be lost, each genotype in the LOD has all the mutations of the previous
one plus exactly one more (as it directly descends from it). The LOD has
no more than one genotype with a given number of mutations, whereas in
general two or more observable genotypes could share the same mutation
count (e.g. in figure \ref{fig:fig1}D both the red and green genotypes
have one mutated driver and are observable). If the evolutionary process
does not conform to SSWM, not all genotypes in the LOD are necessarily
observable. From a clinical perspective, it is relevant to use the
information that is detectable to unveil the underlying dynamics of cancer
evolution, encoded in the LOD. In other words, we want to make a
prediction regarding \textit{what comes next in the LOD} based on the
\textit{observation} of a genotype. We thus formulate the following
general question: \textit{Given that a genotype with $\mathrm{n}$
  mutations has been observed, what genotype with $\mathrm{n}+1$ mutations
  is in the LOD?} Formulated like this, the question is unambiguous and
determined. Under SSWM conditions it can be expressed simply as:
\textit{given that a certain genotype has been observed, what genotype
  comes next in the path to tumor fixation?} However, it is important to
emphasize that even if there is no SSWM the more general question remains
well defined.

% how do we do it?

To address this question, we use a collection of evolutionary simulations
from which we obtain the true paths of tumor progression. For every
simulated process we keep track of % what genotypes represent the majority
% of the population as the tumor progressed (i.e., we list 
the most frequent genotypes. Then for each one of them we ask what
genotype has exactly one more mutation while also being a direct ancestor
of the ultimately fixated clone. Averaging over all simulations we obtain
a set of frequencies which represent the conditional probabilities that we
are interested in. We then examine five CPMs: Mutual Hazard Networks (MHN)
\cite{schill2020}, Conjunctive Bayesian Networks (CBN)
\cite{Gerstung2009,Gerstung2011,montazeri_large-scale_2016}, Oncogenetic
Trees (OT) \cite{Desper1999JCB,Szabo2008}, CAncer PRogression Inference
(CAPRI) \cite{capri_bioinformatics,capri_pnas} and CAncer PRogression
Extraction with Single Edges (CAPRESE) \cite{caprese_2014}, some of which
have more than one variant (giving a total of 13 different methods). We
provide these methods with cross-sectional samples taken from the
evolutionary simulations and compare their predictions with the true
probabilities. We explore the factors that affect performance, focusing on
those factors that depend on the properties of the observed genotype. We
want to understand whether CPMs can be used to predict short-term
evolution; for this question, the key response variable is the best
possible performance over all CPMs considered. Although we do address
method choice as a secondary question, this is not our main objective 
\textit{per se}. We discuss the consequences of our results for the analysis of
cancer data by analyzing 25 cancer data sets.

% This work differs from earlier work \cite{diaz-uriarte2019a,hosseini2019a}
% in that we examine a completely different question: previous work dealt
% with predicting full evolutionary trajectories, but here we examine the
% short-term prediction question \textit{Given that a genotype with
%   $\mathrm{n}$ mutations has been observed, what genotype with
%   $\mathrm{n}+1$ mutations is in the LOD?}, which can colloquially be
% summarized as ``What genotype comes next''. In addition, we add a recent
% new method, Mutual Hazard Networks (MHN) \cite{schill2020}, which is very
% different from the rest of the CPMs considered here. Moreover, we are
% examining genotype-specific factors in addition to global fitness
% landscape characteristics: in contrast to the more general, non-local
% descriptors that may not apply to the totality of the fitness
% landscape/evolutionary process, these genotype-specific factors can
% provide more detailed insight regarding the requirements for the methods
% to be reliable in what are arguably the more relevant questions for
% individualized and precision medicine.

% Why does this matter? Predicting important. In fact,

% experimental evolution studies in bacteria and yeast have allowed to
% partially disentangle the role of population size, selection, and
% mutational heterogeneity in parallel evolution \cite{bailey2017} and,
% thus, indirectly on the issues addressed here; using these results for
% cancer, however, would require much more detailed information (e.g.,
% effective population sizes) than is often available.

% So, what can we do with experimental data from cancer? This we examine in
% the discussion blablaba

% \idea{We probably need to make it explicit how it differs from previous
%   work, in particular Diaz-Uriarte and Vasallo.  If we end up doing what I
% think about comparing JS of methods and using that for real data sets, an
% additional comment is ``and we provide specific, practical advice that can
% be followed with the output of methods applied on real data'' ---or
% something of that sort}.
% \idea{Relate to this: Detection of a specific tumor composition at a given point in
% time automatically discards all mutational pathways that do not visit the
% observed state, reducing the amount of possible evolutionary trajectories
% that can be followed from then on and thus potentially facilitating the
% task of predicting further tumor progression. And relate to Bank}

% \idea{Say something about trying to identify, from method output itself,
%   when good performance}

% \idea{We use MHN. First to review that method. We use CPMs, but also
%   include MHN, a newly developed approach that blablab}

% \idea{This is KEY: we do what really matters: next genotype. It really
%   matters because: a) global predictions suck and are generally doomed
%   except maybe to estimate predictability; b) global predictions can
%   not benefit from local features; c) newxt is what matters for therapy
%   and adaptive treatment}.

\section{Methods} \label{methods}

\subsection{Overview}
\label{sec:overview}

To assess if we can use CPMs for short-term predictions of evolutionary
processes, we have compared predicted and true answers to the question
\textit{Given that a genotype with $\mathrm{n}$ mutations has been
  observed, what genotype with $\mathrm{n}+1$ mutations is in the LOD?}
under an extensive set of simulated scenarios. First, we simulated a large
number of tumor evolution trajectories under scenarios with varying
departures from the assumptions of CPM methods; next, we obtained a large
number of simulated tumor evolution trajectories.  Next, we emulated
cross-sectional sampling of the simulations, with varying sampling schemes
and sample sizes. These steps are described in section
\nameref{methods:simulsandsampling}.  The transition probabilities that
provide the true answer to the question \textit{Given that a genotype with
  $\mathrm{n}$ mutations has been observed, what genotype with
  $\mathrm{n}+1$ mutations is in the LOD?}  is explained in section
\nameref{sec:methods:sim}. To obtain the CPM predictions we analyzed each
of the sampled data sets with a total of 13 different CPM variants and
obtained the predictions as explained in section
\nameref{methods:cpm}. Next, we measured how close the predictions were to
the truth (section \nameref{sec:js}) and, finally, we analyzed the major
factors that affect the similarity between predictions and truth (section
\nameref{sec:linear-mixed-model}).  To examine the implications of our
results for the analyses of cancer data, we have analyzed 25 cancer data
sets, described in section \nameref{sec:datasets}.

\subsection{Simulated evolutionary processes and sampling} \label{methods:simulsandsampling}

Simulated data were taken from \cite{diaz-uriarte2019a}, where complete
details are provided. In summary, simulations were conducted under
evolutionary scenarios that differed in the number of genes, the type of
fitness landscape, the initial population size, and the mutation
rates. Landscapes were of either 7 or 10 genes. Fitness landscapes were of
three types: representable, local maxima or Rough Mount Fuji (RMF). The
three types of fitness landscapes incorporate increasing departures from
the assumptions of most CPMs: for the representable fitness landscapes a
DAG of restrictions exists with the same accessible genotypes and
accessible mutational paths; for local maxima fitness landscapes the set
of accessible genotypes can be represented by a DAG of restrictions, but
there are local fitness maxima and the fitness graph has missing paths
\cite{diaz-uriarte2019a}; the genotype with all genes mutated might or
might not be the genotype with largest fitness. The RMF fitness landscapes
\cite{neidhart_adaptation_2014, devisser_empirical_2014,fragata2019}
usually have multiple local fitness maxima and considerable reciprocal
sign epistasis so not even the set of accessible genotypes can be
represented by a DAG of restrictions. On the above fitness landscapes
evolutionary simulations were run. Three different populations sizes
($2\times10^3$, $5\times10^4$ or $1\times10^6$) and two mutation rate
regimes (mutation rate fixed at $10^{-5}$ or uniformly distributed in the
log scale between $0.2\times10^{-5}$ and $5\times10^{-5}$) were used to
generate variability with respect to SSWM. For each combination of the
above factors, 35 independent replicates were obtained for a total of 1260
landscapes (35 replicates $\times$ 2 number of genes $\times$ 3 types of
fitness landscapes $\times$ 3 initial population sizes $\times$ 2 mutation
rate regimes). Then, for each one of them, 20000 independent evolutionary
processes were simulated until fixation of one of the genotypes at a
fitness maximum (local or global). The complete history of the simulations
was stored so that the true LOD for each simulation was known. Each set of
simulations was then sampled under three detection regimes: with equal
probability with respect to the size distribution (\textit{uniform}
sampling), sampling predominantly early in the tumor progression process
(i.e., enriched in \textit{small} tumors), sampling predominantly late in
the tumor progression process (i.e., enriched in \textit{large}
tumors). When \textit{sampling}, we identified the most abundant genotype
and its mutated loci at the chosen time, so the sampling process is
consistent with how we defined \textit{observable} genotypes earlier. No
sampling error is considered. The full sample consists of three binary
matrices of observations (one per detection regime), each with 20000 rows
(one observation per simulation) and as many columns as driver genes. Each
element of the matrix is either 1 or 0 (true or false) depending on which
genes were mutated in which observations. For each detection regime, the
complete sample was used to generate five smaller, non-overlapping splits
of size 50, 200 or 4000. This produced a total of 56700 data sets (1260
fitness landscapes $\times$ 3 detection regimes $\times$ 3 sample sizes
$\times$ 5 splits).

\subsection{Transition probabilities from evolutionary simulations}
\label{sec:methods:sim}

Let us consider a matrix $\mathbf{P}$ where rows correspond to observable
genotypes and columns to genotypes of the LOD. The element $(i,j)$ of the
matrix, denoted as $p_{ij}$, represents the conditional probability that
the genotype $j$ (with $n+1$ mutations) is an element of the LOD given the
genotype $i$ (with $n$ mutations) has been observed. In the absence of
back mutations and when mutations accumulate one by one, $p_{ij}$ is
defined only if $i$ and $j$ are exactly one mutation away from each other.
Note that the genotype $j$ (element of the LOD and with $n+1$ mutations)
need not be a direct descendant of $i$ if SSWM does not hold. We built a
transition matrix $\mathbf{P}$ for every fitness landscape that we
generated by examining each of the 20000 evolutionary processes that were
simulated in the corresponding landscape, listing all observable genotypes
(i.e. those that represented the majority of the population at some point
during the simulated evolutionary process) and extracting the LOD of every
process tracking back the ancestors of the fixated genotype (definition
and details in \cite{szendro_predictability_2013}; using implementation in
\cite{diaz-uriarte2017}).  Each element $p_{ij}$ can thus be interpreted
as the fraction of the times that $j$ was in the LOD given that an
observation of $i$ had been made. See further details in Supplementary
Material, section \ref{S-supp-cpm-details}.

% \subsubsection{Special cases}\label{sec:special}

% When the end of the evolutionary process is reached there are no further
% transitions to genotypes of the LOD. Thus, matrix $\mathbf{P}$, which
% provides the true probability distributions for the question \textit{what
%   genotype of the LOD has $\mathrm{n}+1$ mutations iven that a genotype
%   with n mutations has been observed?} needs to add to additional states
% that correspond to two distinct special cases. When the observed genotype
% is the final, fixated one, we have simply reached the end of the
% evolutionary process; we handle this by adding to the $\mathbf{P}$ an
% ``end" situation (denoted as $p_{i,end}$). If SSWM conditions are not
% satisfied, however, observable genotypes need not be a part of the LOD so
% that there could be observable genotypes with more mutated loci than, or
% as many mutated loci as, the final genotype that gets fixated. In other
% words, it is possible that a clone with $m \geq n$ mutations establishes
% in the tumor for a period of time even if eventually another clone with
% $n$ mutations is the one that fixates.  We handle this by adding to the
% $\mathbf{P}$ a ``none" situation ($p_{i,none}$). Downstream analyses are
% performed including these two probabilities ($p_{i,end}, p_{i, none}$)
% (and row normalization was actually carried out after including $end$ and
% $none$).

% % \subsection{Transition probabilities from evolutionary simulations}
% % \label{methods:sim}

% % Let us consider a matrix $\mathbf{P}$ where rows correspond to observable
% % genotypes and columns to genotypes of the LOD. The element $(i,j)$ of the
% % matrix, denoted as $p_{ij}$, represents the conditional probability that
% % the genotype $j$ (with $n+1$ mutations) is an element of the LOD given the
% % genotype $i$ (with $n$ mutations) has been observed. Naturally, $p_{ij}$
% % is defined only if $i$ and $j$ are exactly one mutation away from each
% % other. Note that, given how our question is formulated, all the mutations
% % in $i$ need to also be present in $j$ only in SSWM, when $i$ would
% % necessarily be the parent of $j$. But in a more general scenario (e.g.,
% % when SSWM does not hold) it is possible that a genotype $i$ with $n$
% % mutations is observed despite not belonging to the LOD. Therefore the
% % genotype $j$ (element of the LOD and with $n+1$ mutations) need not be a
% % direct descendant of $i$. The only condition for $p_{ij}$ to be defined is
% % that $j$ has exactly one more mutation than $i$:

% % \begin{equation}
% % p_{ij} \in \mathbb{R} \iff n_{mut}(j) = n_{mut}(i) + 1
% % \label{eq:pij}
% % \end{equation} 
% % \idea{En realidad, no es $\mathbb{R}$; es $[0, 1]$, puesto que podemos
% %   definirlo como 0. }

% % We built a matrix $\mathbf{P}$ for every fitness landscape that we
% % generated. To obtain the elements $p_{ij}$ of one of them, we ran through
% % all 20000 evolutionary processes that were simulated in the corresponding
% % landscape. In each case, we first listed all observable genotypes
% % (i.e. those that represented the majority of the population at some point
% % during the simulated evolutionary process). Then we extracted the LOD of
% % every process tracking back the ancestors of the fixated
% % genotype (see definition and details in \cite{szendro_predictability_2013}; using implementation in \cite{diaz-uriarte2017}).
% % We identified the genotypes in the LOD with one mutation more than the
% % observable ones, and added 1 to the corresponding entries of
% % $\mathbf{P}$. Finally, we row-normalized $\mathbf{P}$ to meet the
% % normalization condition $\sum_{j} p_{ij} = 1$.

% % % \begin{equation}
% % % \sum_{j} p_{ij} = 1
% % % \label{eq:pij-norm}
% % % \end{equation}\todo{Added this equation, OK?}
% % %
% % Each element $p_{ij}$ can thus be interpreted as the fraction of the times that $j$ was in the LOD given that an observation of $i$ had been made. For a large number of simulations, this fraction represents a good approximation of the true probabilities.

% % \subsubsection{Special cases}\label{sec:special}

% % If SSWM conditions are not satisfied, observable genotypes need not be a part of the LOD and not all genotypes in the LOD are necessarily observable. A particular case of this are observable genotypes with more mutated loci (or just as many) than the final one that gets fixated: it is possible that a clone with $m \geq n$ mutations establishes in the tumor for a period of time even if eventually another clone with $n$ mutations is the one that fixates. This represents a special scenario for our purpose: the answer to out question (\textit{what genotype of the LOD has $\mathrm{n}+1$ mutations?}) is technically undefined if the LOD stops at $n$ mutations. We identify two scenarios: \textit{a)} the observed genotype is simply the final, fixated one, and \textit{b)} the observed genotype has transiently established as the most abundant, but fixation will occur at a different one (with less mutations) evolving through a different pathway.
% % \iffalse

>>>>

\documentclass[a4paper,10pt]{article}

% packages
\usepackage[utf8]{inputenc} % allows usage of spanish special characters
\usepackage[spanish,english]{babel} % english dictionary for proper hyphenation
\usepackage{amsmath} % math expressions
\usepackage{upgreek} % upright greek letters for math
\usepackage{txfonts} % nice fonts 
\usepackage{authblk} % author affiliations
\usepackage{graphicx} % images
\usepackage{float} % image positioning
\usepackage{multirow} % allows merging cells in tables
\usepackage{makecell} % more table customization
\usepackage{mathtools} % nice matrices
\usepackage{rotating} % rotated text
\usepackage{caption} % customization of figure/table captions
\usepackage{subcaption} % collages of multiple images
\usepackage{hyperref} % hyperlinks
\usepackage{nameref} % cross-references between draft and supplementary material
\usepackage{zref-xr,zref-user} % more cross-references formatting
\usepackage[square,numbers,sort&compress]{natbib}
\usepackage[table,xcdraw]{xcolor} % text and table background colors
\usepackage[margin=3cm]{geometry} % margins
\usepackage[nottoc,numbib]{tocbibind} % add references to table of contents
\usepackage{lineno} % line numbers

\usepackage{pdflscape} %% Landscape
\usepackage{dcolumn}
\usepackage{booktabs}
\usepackage{savesym}
\savesymbol{iint}
\restoresymbol{TXF}{iint}
%% \usepackage{pdfpages} %% to include a pdf file

% customization
\setcounter{Maxaffil}{0} % affiliations
\renewcommand\Affilfont{\itshape\small} % style of affiliations text
\makeatletter\renewcommand{\@biblabel}[1]{#1.}\makeatother % change [number] for number. in reference list
\addto{\captionsenglish}{\renewcommand{\bibname}{References}}
\newcommand{\idea}[1]{\textcolor{red}{#1}}
\newcommand{\mr}{\textcolor{red}{\textbf{[missing ref(s)]}}}
\newcommand{\Burl}[1]{\url{#1}}
\zxrsetup{toltxlabel} % cross references: draft & suppl mat
\zexternaldocument*[M-]{draft}
\hypersetup{
  colorlinks = true,
  citecolor=  black,
  linkcolor = {blue},
  filecolor = cyan % controls color of external ref, if used
}
\renewcommand{\thefigure}{\textbf{S\arabic{figure}}} % preceed the Suppl Mat figures with S
\renewcommand{\thetable}{\textbf{S\arabic{table}}} % preceed the Suppl Mat tables with S
\renewcommand{\theequation}{S\arabic{equation}} % preceed the Suppl Mat equations with S
\captionsetup[figure]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Fig.}} % custom figure captions
\captionsetup[table]{font=small, labelfont={bf}, labelformat={default},
   labelsep=period, name={Table}} % custom table captions
\hypersetup{citecolor={blue}}
\definecolor{lightgray}{gray}{0.9}

% title, authors, affiliations
\title{Provisional Title}
\author[1,2,3]{Authors}
\author[1,2,$\dagger$]{Álvaro Sánchez}
\affil[1]{Department of Ecology \& Evolutionary Biology,
Yale University, New Haven, CT, USA}
\affil[2]{Microbial Sciences Institute,
Yale University, New Haven, CT, USA}
\affil[3]{Other affiliations...}
\affil[$\dagger$]{To whom correspondence should be addressed: \normalfont alvaro.sanchez@yale.edu}
\date{}

\newcommand\legendtreefigs{Internal nodes show the partitioning variables   and the values that lead to each partition; light blue denotes the   values or categories that lead down the left split and salmon the values   or categories that lead down the right split (partitioning criteria are   also shown in the edges leading down a node). Leaves or terminal nodes   show (weighted) boxplots of the dependent variable for the observations   that belong that fall in that leave and, as label, the name of the node   ax_3,nd the sample size (sum of weighted observations and \% of the total   data set for the fit). The dot plots underneath the boxplots show the mean JS for all the methods for that leave.}

@@

\linenumbers

@@

The abstract goes here.

@@

\section{Introduction}\label{intro}

This is an example cite \cite{Vetrovsky2013,nloptr}
